<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{An R Markdown Vignette made with knitr}
-->


fingis - tutorial
===========

This R package provides tools to access open spatial data in Finland
as part of the [rOpenGov](http://ropengov.github.io) project.

For contact information and source code, see the [github page](https://github.com/rOpenGov/fingis)

## Available data sources

The following data sources are currently available:
* [Helsinki region district maps](#aluejakokartat) (Helsingin seudun aluejakokartat)
 * Aluejakokartat: kunta, pien-, suur-, tilastoalueet (Helsinki region district maps)
 * Äänestysaluejako: (Helsinki region election district maps)
 * Source: [Helsingin kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/)
* [Helsinki spatial data](#hel-spatial) (Helsingin seudun avoimia paikkatietoaineistoja)
 * Seutukartta (Helsinki Region Maps)
 * Helsingin piirijako (District Division of the City of Helsinki)
 * Rakennusrekisterin ote (Helsinki building registry)
 * Source: [Helsingin kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/)
* [National Land Survey data](#maanmittauslaitos) (Maanmittauslaitoksen avointa dataa)
 * Yleiskartat: kunta-, maakuntarajat (municipality and province borders)
 * Source: [Maanmittauslaitos (MML)](http://www.maanmittauslaitos.fi/avoindata)
* [Geocoding](#geocoding)
 * [OKF.fi Geocoding API Test Console](http://data.okf.fi/console/)
 * [OpenStreetMap Nominatim](http://wiki.openstreetmap.org/wiki/Nominatim_usage_policy)
 * [Google](http://code.google.com/apis/maps/documentation/geocoding/)

## Installation

Release version for general users:

```{r install, eval=FALSE}
install.packages("fingis")
```

Development version for developers:

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("fingis", "ropengov")
```

Load package.

```{r load}
library(fingis)
```

### Notes

The fingis package uses the [rgdal](http://cran.r-project.org/web/packages/rgdal/index.html) library, which depends on the [GDAL](http://www.gdal.org/) spatial framework. Some rgdal installation tips for various platforms lister below. If you encounter problems, please contact us by email: louhos@googlegroups.com.
* Windows: Install binaries from [CRAN](http://cran.r-project.org/web/packages/rgdal/index.html)
* OSX: Install binaries from [CRAN](http://cran.r-project.org/web/packages/rgdal/index.html). Check also [KyngChaos Wiki](http://www.kyngchaos.com/software/frameworks) 
* Linux: Try the installation scripts [here](https://github.com/louhos/takomo/tree/master/installation/) (not necessarily up-to-date!)

## <a name="aluejakokartat"></a>Helsinki region district maps

Helsinki region district maps (Helsingin seudun aluejakokartat) from [Helsingin kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/).

List available maps with `get_helsinki_aluejakokartat()`.

```{r aluejako_list, message=FALSE}
get_helsinki_aluejakokartat()
```

Below the 'suuralue' districts is used for plotting examples with `spplot()` and [ggplot2](http://ggplot2.org/). The other district types can be plotted similarly.

### Plot with spplot

Retrieve 'suuralue' spatial object with `get_helsinki_aluejakokartat()` and plot with `spplot()`.

```{r hkk-suuralue1, message=FALSE, warning=FALSE, results='hide'}
sp.suuralue <- get_helsinki_aluejakokartat(map.specifier="suuralue")
spplot(sp.suuralue, zcol="Name")
```

Function `generate_map_colours()` allows nice region colouring separable adjacent regions. This is used here with the `rainbow()` colour scale to plot the regions with `spplot()`.

```{r hkk-suuralue2, message=FALSE, warning=FALSE, results='hide'}
sp.suuralue@data$COL <- factor(generate_map_colours(sp=sp.suuralue))
spplot(sp.suuralue, zcol="COL", col.regions=rainbow(length(levels(sp.suuralue@data$COL))), colorkey=FALSE)
```

### Plot with ggplot2

Use the 'sp.suuralue' retrieved above, and retrieve also the center points of the districts. Use `sp2df()` function to tranform the spatial objects into data frames. Plot with [ggplot2](http://ggplot2.org/), using blank map theme with `get_theme_map()`. 

```{r hkk-suuralue3, message=FALSE, warning=FALSE, results='hide'}
# Retrieve center points
sp.suuralue.piste <- get_helsinki_aluejakokartat(map.specifier="suuralue_piste")
# Get data frames
df.suuralue <- sp2df(sp.suuralue, "Name")
df.suuralue.piste <- sp2df(sp.suuralue.piste, "Name")
# Set map theme
library(ggplot2)
theme_set(get_theme_map())
# Plot regions, add labels using the points data
ggplot(df.suuralue, aes(x=long, y=lat)) + geom_polygon(aes(fill=COL, group=Name)) + geom_text(data=df.suuralue.piste, aes(label=Name)) + theme(legend.position="none")
```

Add background map from OpenStreetMap using `get_map()` from [ggmap](https://sites.google.com/site/davidkahle/ggmap) and plot again.

```{r hkk-suuralue4, message=FALSE, warning=FALSE, results='hide'}
# Add background map from OpenStreetMap using ggmap
library(ggmap)
# Get bounding box from sp.suuralue
hel.bbox <- as.vector(sp.suuralue@bbox)
# Get map using openstreetmap
hel.map <- ggmap::get_map(location=hel.bbox, source="osm")
# Plot transparent districts on top the background map
ggmap(hel.map) + geom_polygon(data=df.suuralue, aes(x=long, y=lat, fill=COL, group=Name), alpha=0.5) + geom_text(data=df.suuralue.piste, aes(x=long, y=lat, label=Name)) + theme(legend.position="none")
```

### Plot election districts

Retrieve and plot äänetysaluejako (election districts) with `get_helsinki_aluejakokartat()` and `spplot()`, use colours to separate municipalities.

```{r hkk-aanestysalue, message=FALSE, warning=FALSE, results='hide'}
sp.aanestys <- get_helsinki_aluejakokartat(map.specifier="aanestysalue")
spplot(sp.aanestys, zcol="KUNTA", col.regions=rainbow(length(levels(sp.aanestys@data$KUNTA))), colorkey=FALSE)

```


## <a name="hel-spatial"></a>Helsinki spatial data

Other Helsinki region spatial data from [Helsingin Kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/).

List available spatial data with `get_helsinki_spatial()`.

```{r hki_spatial_list, message=FALSE}
get_helsinki_spatial()
```


## <a name="maanmittauslaitos"></a>National Land Survey Finland

Spatial data from [National Land Survey Finland](http://www.maanmittauslaitos.fi/en/opendata)  (Maanmittauslaitos, MML). These data are preprocessed into RData format, see details [here](https://github.com/avoindata/mml).

List available data sets with `list_mml_datasets()`.

```{r MML_list, message=FALSE} 
list_mml_datasets()
```

Retrieve regional borders for Finland with `get_mml()`.

```{r MML_regions, message=FALSE, warning=FALSE} 
# Get a specific map
sp.mml <- get_mml(map.id="Yleiskartta-4500", data.id="HallintoAlue")
# Investigate available variables in this map
head(as.data.frame(sp.mml))
```

Plot municipalities (kunnat) with `spplot()`, using colours from `generate_map_colours()`.


```{r MML_province, message=FALSE, warning=FALSE}
# Get region colouring for municipalities
sp.mml@data$COL <- factor(generate_map_colours(sp.mml))
# Plot the shape file, colour municipalities
spplot(sp.mml, zcol="COL", col.regions=rainbow(length(levels(sp.mml@data$COL))), colorkey=FALSE)
```

Plot provinces (maakunnat) with `spplot()`. Note that `generate_map_colours()` works currently only for the smallest polygons in the data, here municipalities, and can not be used for the provinces!

```{r MML_municipality, message=FALSE, warning=FALSE}
# Convert municipality names to factors and fix character encoding
sp.mml@data$Maakunta.FI <- factor(iconv(sp.mml@data$Maakunta.FI, from="UTF-8", to="ISO-8859-1"))
# Plot the shape file, colour provinces 
spplot(sp.mml, zcol="Maakunta.FI", col.regions=rainbow(length(levels(sp.mml@data$Maakunta.FI))))
```

## <a name="geocoding"></a>Geocoding

Get geocodes for given location (address etc.) using one of the available services. Please read carefully the usage policies for the different services:
* [OKF.fi Geocoding API Test Console](http://data.okf.fi/console/)
* [OpenStreetMap Nominatim](http://wiki.openstreetmap.org/wiki/Nominatim_usage_policy)
* [Google](http://code.google.com/apis/maps/documentation/geocoding/)

The function `get_geocode()` returns both latitude and longitude for the first hit, and the raw output (varies depending on the service used).

Warning! The geocode results may vary between sources, use with care!


```{r geocoding, message=FALSE}
gc1 <- get_geocode("Mannerheimintie 100, Helsinki", service="okf")
unlist(gc1[1:2])
gc2 <- get_geocode("Mannerheimintie 100, Helsinki", service="openstreetmap")
unlist(gc2[1:2])
gc3 <- get_geocode("Mannerheimintie 100, Helsinki", service="google")
unlist(gc3[1:2])
```

### Citation

**Citing the data:** See `help()` to get citation information for each data source individually.

**Citing the R package:**

```{r citation, comment=NA}
citation("fingis")
```


### Session info


This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```

