<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{An R Markdown Vignette made with knitr}
-->

Finland GIS R tools
===========

This is an [rOpenGov](https://github.com/rOpenGov/fingis) R package
providing tools for Finnish GIS data.

## Availablae data

The following data sets are currently available:
* [Installation](#installation)
* [Helsinki region aluejakokartat](https://github.com/rOpenGov/fingis/#helsinki-region-aluejakokartat)
* [Maanmittauslaitos](https://github.com/rOpenGov/fingis/#maanmittauslaitos)
* [Geocoding](https://github.com/rOpenGov/fingis/#geocoding)

### Installation

Note! The fingis package uses the rgdal library, which depends on the GDAL spatial framework. Installing this might be tricky. 

FIXME: Add rgdal installation tips for all platforms:
* Windows:
* Linux: 
* OSX: Check [KyngChaos Wiki](http://www.kyngchaos.com/software/frameworks) 

Release version for general users (NOT AVAILABLE YET):


Development version for developers:

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("fingis", "ropengov")
library(fingis)
```

FIXME: link below

Further installation and development instructions at the [home
page](https://github.com/rOpenGov/fingis).

### Helsinki region aluejakokartat

Retrieve Helsinki region aluejakokartat (district maps.)

List available maps

```{r aluejako_list, message=FALSE}
map.list <- get_Helsinki_aluejakokartat()
map.list
```

Retrieve 'suuralue' spatial object and plot it with the plot_shape() function.

```{r hkk-suuralue1, message=FALSE, warning=FALSE, results='hide'}
sp.suuralue <- get_Helsinki_aluejakokartat(map.specifier="suuralue")
plot_shape(sp=sp.suuralue, varname="Name", type="discrete", plot=FALSE)
```

Retrieval 'suuralue_piste' spatial object, containing the center points of the districts, and plot with spplot().

```{r hkk-suuralue2, message=FALSE, warning=FALSE, results='hide'}
sp.suuralue.piste <- get_Helsinki_aluejakokartat(map.specifier="suuralue_piste")
sp::spplot(obj=sp.suuralue.piste, zcol="Name")
```

Use the sp2df() function to tranform the spatial objects into data frames. Plot with ggplot2, using blank theme with get_theme_map(). 

```{r hkk-suuralue3, message=FALSE}
# Get data frames
df.suuralue <- sp2df(sp.suuralue, "Name")
df.suuralue.piste <- sp2df(sp.suuralue.piste, "Name")

# Set map theme
theme_set(get_theme_map())
# Plot regions, add labels using the points data
ggplot(df.suuralue, aes(x=long, y=lat, fill=Name)) + geom_polygon() + geom_text(data=df.suuralue.piste, aes(label=Name)) + theme(legend.position="none")
```

Add background map from OpenStreetMap using the [ggmap](https://sites.google.com/site/davidkahle/ggmap) package.

```{r hkk-suuralue4, message=FALSE}
# Add background map from OpenStreetMap using ggmap
library(ggmap)
# Get bounding box from sp.suuralue
hel.bbox <- as.vector(sp.suuralue@bbox)
# Get map using openstreetmap
hel.map <- ggmap::get_map(location=hel.bbox, source="osm")
# Plot transparent districts on top the background map
ggmap(hel.map) + geom_polygon(data=df.suuralue, aes(x=long, y=lat, fill=Name), alpha=0.5) + geom_text(data=df.suuralue.piste, aes(x=long, y=lat, label=Name)) + theme(legend.position="none")
```

Retrieve and plot äänetysaluejako (election districts).

```{r hkk-aanestysalue, message=FALSE, warning=FALSE, results='hide'}
sp.aanestys <- get_Helsinki_aluejakokartat(map.specifier="aanestysalue")
plot_shape(sp.aanestys, "KUNTA", type="discrete", plot=FALSE)
```

### Other Helsinki region spatial data

To be added

### Maanmittauslaitos

Spatial data from  Maanmittauslaitos (MML, Land Survey Finland). These data are preprocessed into RData format (FIXME: add description).

List available data sets

```{r MML_list, message=FALSE} 
list_mml_datasets()
```

Retrieve regional borders for Finland.

```{r MML_regions, message=FALSE, warning=FALSE} 
# Get a specific map
sp.mml <- get_MML(map.id="Yleiskartta-4500", data.id="HallintoAlue") 

# Investigate available variables in this map
head(as.data.frame(sp.mun))
```

Plot provinces (maakunnat).

```{r MML_province, message=FALSE, warning=FALSE} 
# Plot the shape file, colour provinces
plot_shape(sp=sp.mml, varname="Maakunta", type="discrete", plot = FALSE)
```

Plot municipalities (kunnat).

```{r MML_municipality, message=FALSE, warning=FALSE} 
# Plot the shape file, colour municipalities
plot_shape(sp=sp.mml, varname="Kunta", type="discrete", plot = FALSE)
```

### Geocoding

Get geocodes for given location (address etc.) using one of the available services. Please read carefully the usage policies for the different services:
* [OKF.fi Geocoding API Test Console](http://data.okf.fi/console/)
* [OpenStreetMap Nominatim](http://wiki.openstreetmap.org/wiki/Nominatim_usage_policy)
* [Google](http://code.google.com/apis/maps/documentation/geocoding/)

The function get_geocode() returns both latitude and longitude for the first hit, and the raw output (varies depending on the service used).

```{r geocoding, message=FALSE}
gc1 <- get_geocode("Mannerheimintie 100, Helsinki", service="okf")
unlist(gc1[1:2])
gc2 <- get_geocode("Mannerheimintie 100, Helsinki", service="openstreetmap")
unlist(gc2[1:2])
gc3 <- get_geocode("Mannerheimintie 100, Helsinki", service="google")
unlist(gc3[1:2])
```

### Licensing and Citations

This work can be freely used, modified and distributed under the
[Two-clause FreeBSD
license](http://en.wikipedia.org/wiki/BSD\_licenses). Cite this R
package and and the appropriate data provider, including a url
link. Kindly cite the R package as 'Leo Lahti, Juuso Parkkinen ja
Joona Lehtomäki (2014). fingis R package. URL:
https://github.com/rOpenGov/fingis'.


### Session info


This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```

