---
title: "geofi tutorial"
author: "Markus Kainu, Leo Lahti & Joona Lehtomäki"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{geofi tutorial}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Installation

`geofi` is not yet in CRAN, but can be installed from Github using

```{r, eval = FALSE}
remotes::install_github("ropengov/geofi")
```


## Municipalities

```{r municipality_map, fig.height = 4, fig.width = 7, warning=FALSE}
library(geofi)
library(ggplot2)
library(dplyr)
library(janitor)
library(tidyr)

municipalities <- get_municipalities(year = 2020, scale = 4500)
ggplot(municipalities) + 
  geom_sf(aes(fill = as.integer(kunta)))
```

## Zipcodes


```{r zipcode_map, fig.height = 4, fig.width = 7}
zipcodes <- get_zipcodes(year = 2020) 
ggplot(zipcodes) + 
  geom_sf(aes(fill = as.integer(posti_alue)))
``` 


You can add variables from Paavo data (Statistics Finland) in following manner.

```{r zipcode_with_statistics_finland, fig.height = 4, fig.width = 7}
library(pxweb)
# lets get all zipcodes and all variables
pxweb_query_list <- 
  list("Postinumeroalue"=c("*"),
                           "Tiedot"=c("*"))

# Download data 
px_data <- 
 pxweb_get(url = "http://pxnet2.stat.fi/PXWeb/api/v1/fi/Postinumeroalueittainen_avoin_tieto/2019/paavo_1_he_2019.px",
           query = pxweb_query_list)

# Convert to data.frame 
tk_data_zip <- as.data.frame(px_data, column.name.type = "text", variable.value.type = "text")

library(dplyr)
library(tidyr)
tk_data_zip2 <- tk_data_zip %>%
  as_tibble() %>% 
  setNames(c("region","indicator","value")) %>% 
  spread(indicator, value)
tk_zipdata <- janitor::clean_names(tk_data_zip2)
tk_zipdata$zipcode <- substr(x = tk_zipdata$region, start = 1, stop = 5)
tk_zipdata$zipname <- gsub("^.{6}", "", tk_zipdata$region)


# Lets join with spatial data and plot the area of each zipcode
zipcodes19 <- get_zipcodes(year = 2019) 
zipcodes_new <- left_join(zipcodes19, tk_zipdata, by = c("posti_alue" = "zipcode"))
ggplot(zipcodes_new) + 
  geom_sf(aes(fill = pinta_ala), color = alpha("white", 1/3))
```





## Joining municipality level data from Statistics Finland


```{r municipalities_with_data}
municipalities17 <- get_municipalities(year = 2017)

# pull municipality data from Statistics Finland
library(pxweb)
pxweb_query_list <-
  list("Alue 2019"=c("*"),
       "Tiedot"=c("M408","M411","M476","M391","M421","M478","M404","M410","M303","M297","M302","M44","M62","M70","M488","M486","M137","M140","M130","M162","M78","M485","M152","M72","M84","M106","M151","M499","M496","M495","M497","M498"),
       "Vuosi"=c("2017"))

px_data <-
  pxweb_get(url = "http://pxnet2.stat.fi/PXWeb/api/v1/fi/Kuntien_avainluvut/2019/kuntien_avainluvut_2019_aikasarja.px",
            query = pxweb_query_list)

# Convert to data.frame
tk_data <- as.data.frame(px_data, column.name.type = "text", variable.value.type = "text")
tk_data2 <- tk_data %>%
  rename(name = `Alue 2019`) %>%
  mutate(name = as.character(name),
         # Paste Tiedot and Vuosi
         Tiedot = paste(Tiedot, Vuosi)) %>%
  select(-Vuosi) %>%
  spread(Tiedot, `Kuntien avainluvut`) %>%
  as_tibble()
tk_data3 <- janitor::clean_names(tk_data2)

# Join with Statistics Finland attribute data
dat <- left_join(municipalities17, tk_data3)
dat[1:10,1:10]
```


## Aggregating municipality level data using internal municipality_key-files

```{r municipalities_map, fig.height = 4, fig.width = 7}
# Lets aggregate at maakunta level using column mk_name
mk_data <- dat %>% 
  group_by(maakunta_name_fi,maakunta_code) %>% 
  # And compute population at maakunta level
  summarise(vakiluku_2017 = sum(vakiluku_2017))

ggplot(mk_data) + 
  geom_sf(aes(fill = vakiluku_2017)) +
  geom_sf_text(aes(label = paste(maakunta_name_fi, "\n",vakiluku_2017)), 
               size = 3, 
               color = "white")

```


Or if you prefer aggregating at seutukunta-level you can

```{r vaalipiiri_map, fig.height = 4, fig.width = 7}
mk_data <- dat %>% 
  group_by(seutukunta_code,seutukunta_name_fi) %>% 
  # And compute population at maakunta level
  summarise(vakiluku_2017 = sum(vakiluku_2017))

ggplot(mk_data) + 
  geom_sf(aes(fill = vakiluku_2017)) +
  geom_sf_text(aes(label = paste(seutukunta_name_fi, "\n",vakiluku_2017)), 
               size = 3, 
               color = "white")

```


Or if you want the population at the national level aggregated up from municipality level you can: 

```{r, national_map, fig.height = 4, fig.width = 7}
mk_data <- dat %>% 
  summarise(vakiluku_2017 = sum(vakiluku_2017))

ggplot(mk_data) + 
  geom_sf(aes(fill = vakiluku_2017)) +
  geom_sf_text(aes(label = vakiluku_2017), 
               size = 3, 
               color = "white")

```


## Interactive maps using `leaflet`-package



```{r leaflet, fig.height = 4, fig.width = 7}
library(leaflet)
library(geofi)
library(dplyr)

municipalities <- get_municipalities(year = 2019)
municipalities_lonlat <- sf::st_transform(x = municipalities, crs = "+proj=longlat +datum=WGS84")

leaflet(municipalities_lonlat) %>% 
  addTiles() %>% 
  addPolygons(color = "coral", 
              weight = 1,
              smoothFactor = 0.5,
              opacity = 1.0,
              fillOpacity = 0.5,
    label = ~nimi,
    highlightOptions = highlightOptions(color = "white", weight = 2,
      bringToFront = TRUE))

```

# Population grid data

## Whole country at 5km by 5km

```{r, population_grid_data, fig.height = 4, fig.width = 7}
pop_grid <- get_population_grid(year = 2018, resolution = 5)
ggplot(pop_grid) + 
  geom_sf(aes(fill = vaesto), color = NA) +
  scale_fill_viridis_c(trans = "sqrt", option = "plasma")
```

## Helsinki at 1km by 1km as interactive leaflet map

```{r, population_grid_data2, fig.height = 4, fig.width = 7}
pop_grid <- get_population_grid(year = 2018, resolution = 1)

pop_grid_helsinki <- sf::st_transform(x = pop_grid %>% 
                                        filter(kunta == "091"), crs = "+proj=longlat +datum=WGS84")



leaflet(pop_grid_helsinki) %>% 
  addTiles() %>% 
  addPolygons(fillColor = ~colorQuantile("RdYlGn", vaesto)(vaesto),
              color = "coral", 
              weight = 1,
              smoothFactor = 0.5,
              opacity = 1.0,
              fillOpacity = 0.5,
    label = ~vaesto,
    highlightOptions = highlightOptions(color = "white", weight = 2,
      bringToFront = TRUE))
```


# Using custom geofacet grids

From Ryan Hafen's [blog](https://ryanhafen.com/blog/geofacet/):

>The [geofacet](https://hafen.github.io/geofacet/) package extends [ggplot2](http://ggplot2.tidyverse.org/) in a way that makes it easy to create geographically faceted visualizations in R. To geofacet is to take data representing different geographic entities and apply a visualization method to the data for each entity, with the resulting set of visualizations being laid out in a grid that mimics the original geographic topology as closely as possible.

`geofi`-package contains custom grids to be used with various Finnish administrative breakdowns. Below is an example where population data at municipality level is pulled from Statistics Finland fom 1987 to 2018, then aggregated at the levels of regions (maakunta) and then plotted with ggplot2 using grid `geofi::grid_mk_2019`.


```{r geofacet, fig.height = 6, fig.width = 7}

# Let pull population data from Statistics Finland
pxweb_query_list <- 
  list("Alue 2019"=c("*"),
       "Tiedot"=c("M411"),
       "Vuosi"=c("1987","1988","1989","1990","1991","1992","1993","1994","1995","1996","1997","1998","1999","2000","2001","2002","2003","2004","2005","2006","2007","2008","2009","2010","2011","2012","2013","2014","2015","2016","2017","2018"))

# Download data 
px_data <- 
  pxweb_get(url = "http://pxnet2.stat.fi/PXWeb/api/v1/fi/Kuntien_avainluvut/2019/kuntien_avainluvut_2019_aikasarja.px",
            query = pxweb_query_list)

# Convert to data.frame 
px_data <- as.data.frame(px_data, column.name.type = "text", variable.value.type = "text")
names(px_data) <- c("kunta_name","var","year","value")

# lets aggregate population data
dat <- left_join(geofi::municipality_key_2019 %>% select(-year),
                 px_data) %>% 
  group_by(maakunta_code, maakunta_name_fi,year) %>% 
  summarise(population = sum(value, na.rm = TRUE)) %>% 
  na.omit() %>% 
  ungroup() %>% 
  rename(code = maakunta_code, name = maakunta_name_fi)

library(geofacet)
library(ggplot2)

ggplot(dat, aes(x = year, y = population/1000, group = name)) + 
  geom_line() + 
  facet_geo(facets = ~name, grid = grid_maakunta_2019, scales = "free_y") +
  theme(axis.text.x = element_text(size = 6)) +
  scale_x_discrete(breaks = seq.int(from = 1987, to = 2018, by = 5)) +
  labs(title = "Population 1987-2018", y = "population (1000)")
```


```{r geofacet2, fig.height = 6, fig.width = 7}

dat <- left_join(geofi::municipality_key_2019 %>% 
                   filter(maakunta_name_fi == "Ahvenanmaa") %>% 
                   select(-year),
                 px_data) %>% 
  group_by(maakunta_code, maakunta_name_fi,year) %>% 
  rename(code = kunta, name = kunta_name, population = value)

library(geofacet)
library(ggplot2)

ggplot(dat, aes(x = year, y = population/1000, group = name)) + 
  geom_line() + 
  facet_geo(facets = ~name, grid = grid_ahvenanmaa_2019, scales = "free_y") +
  theme(axis.text.x = element_text(size = 6)) +
  scale_x_discrete(breaks = seq.int(from = 1987, to = 2018, by = 5)) +
  labs(title = "Population 1987-2018", y = "population (1000)")
```

