---
title: "Accessing Spatial Data from NLS OGC apis with geofi"
author: "Markus Kainu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accessing Spatial Data with geofi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
library(geofi)
library(sf)
library(dplyr)
library(ggplot2)
```

## Introduction

The `geofi` package provides tools to access spatial data from the **Maastotietokanta** (Topographic Database) and **Geographic Names** (Nimistö) datasets provided by the National Land Survey of Finland (NLS) via their OGC API. This vignette demonstrates how to use the package's core functions to:

* List available collections in the Maastotietokanta.
* Download specific collections of spatial data.
* Query geographic place names with flexible filtering options.

The package handles API authentication, pagination, spatial filtering, and coordinate reference system (CRS) transformations, making it easy to work with Finnish spatial data in R.

To use the package, you need an **API key** from the National Land Survey of Finland. Visit [their website](https://www.maanmittauslaitos.fi/en/rajapinnat/api-avaimen-ohje) to obtain one. Once you have the key, set it in R using:


```{r, eval = FALSE}
options(geofi_mml_api_key = "your_api_key_here")
```


```{r, include = FALSE}
options(geofi_mml_api_key = Sys.getenv("MML_API_KEY"))
```



## Package Overview

The `geofi` package includes the following key functions:

* `ogc_get_maastotietokanta_collections()`: Retrieves a list of available collections from the Maastotietokanta, including their titles and descriptions.
* `ogc_get_maastotietokanta()`: Downloads a specific collection from the Maastotietokanta, with options for CRS, spatial filtering, and pagination.
* `ogc_get_nimisto()`: Queries the Geographic Names dataset, allowing filtering by search string, bounding box, and other parameters.
* `fetch_ogc_api_mml()`: An internal function that handles low-level API requests and pagination (not typically called directly by users).

All functions return spatial data as `sf` objects, compatible with the `sf` package for spatial analysis and visualization.

## Step 1: Listing Available Collections

To explore the datasets available in the Maastotietokanta, use `ogc_get_maastotietokanta_collections()`. This function queries the OGC API and returns a data frame with collection IDs and descriptions.

```{r}
collections <- ogc_get_maastotietokanta_collections()
head(collections)
```

This list helps you identify the `collection` parameter needed for `ogc_get_maastotietokanta()`.

## Step 2: Downloading a Maastotietokanta Collection

The `ogc_get_maastotietokanta()` function downloads a specific collection as an `sf` object. You can customize the output with parameters like:

* `collection`: The name of the collection (e.g., "`hautausmaa`" for cemeteries).
* `crs`: The coordinate reference system (EPSG:3067 or EPSG:4326).
* `limit`: Maximum number of features per request (or `NULL` to fetch all).
* `bbox`: A bounding box for spatial filtering (in EPSG:4326).
* `max_pages`: Maximum number of pages to fetch when paginating.

### Example: Downloading Cemeteries

Let's download the "`hautausmaa`" collection in the default CRS (EPSG:3067):

```{r}
cemeteries <- ogc_get_maastotietokanta(collection = "hautausmaa", crs = 4326)
cemeteries
```

### Example: Spatial Filtering with a Bounding Box

To download cemeteries within a specific area (e.g., around Helsinki), use the `bbox` parameter. Coordinates must be in EPSG:4326 (WGS84).

```{r}
cemeteries_helsinki <- ogc_get_maastotietokanta(
  collection = "hautausmaa",
  bbox = "24.5,60.1,25.5,60.5",
  crs = 4326
)
```

Visualize the results using `ggplot2`:

```{r}
ggplot(cemeteries_helsinki) +
  geom_sf() +
  theme_minimal() +
  labs(title = "Cemeteries near Helsinki")
```

### Example: Handling Large Collections

For large collections like "`suo`" (bogs/marshes), you may need to increase `max_pages` to fetch all features:

```{r, eval = FALSE}
bogs <- ogc_get_maastotietokanta(
  collection = "suo",
  max_pages = 15
)
```

**Note**: Large collections may take significant time to download. If `max_pages` is reached, a warning will indicate that additional features may exist.

## Step 3: Querying Geographic Names

The `ogc_get_nimisto()` function retrieves place names from the Geographic Names dataset. You can filter results by:

* `search_string`: A case-insensitive search term (e.g., "`kainu`").
* `bbox`: A bounding box for spatial filtering.
* `limit`: Maximum number of features to retrieve.
* `crs`: Output CRS (EPSG:3067 or EPSG:4326).

### Example: Searching for Place Names

Search for place names containing "`kainu`":

```{r}
kainu_places <- ogc_get_nimisto(search_string = "kainu")
print(kainu_places)
```

Visualize the results:

```{r}
ggplot(kainu_places) +
  geom_sf() +
  geom_sf_text(aes(label = spelling), size = 3, check_overlap = TRUE) +
  theme_minimal() +
  labs(title = "Place Names Containing 'Kainu'")
```

### Example: Combining Search and Spatial Filtering

Search for "kainu\*" (with \* wildcard) within a bounding box covering most part of Kainuu-region:

```{r}
kainu_bbox <- ogc_get_nimisto(
  search_string = "kainu*",
  bbox = "27.515259,63.450509,30.531006,64.524823",
  crs = 4326
)
```

## Advanced Features

### Pagination

When `limit = NULL` in `ogc_get_maastotietokanta()` or `ogc_get_nimisto()`, the package automatically paginates through large datasets. The `fetch_ogc_api_mml()` function handles this internally, attempting to fetch all features in one request (`limit=-1`) or falling back to paginated requests if needed. Use `max_pages` to control the maximum number of pages fetched.

### Error Handling

The package includes robust error handling:

* Validates inputs (e.g., API key, CRS, bounding box format).
* Retries failed requests up to 3 times for transient network issues.
* Handles HTTP 429 (rate limit) errors by respecting the Retry-After header.
* Provides informative error messages for API failures or invalid responses.

### Custom Parameters

Function `ogc_get_nimisto()` support `custom_params` for advanced API queries. For example:

```{r}
swimming_beaches_filtered_by_municipality_number <- ogc_get_nimisto(
  search_string = "*uimaranta*",
  custom_params = "municipality=091"
)
swimming_beaches_filtered_by_municipality_number
```

Consult the [NLS OGC API documentation](https://www.maanmittauslaitos.fi/nimiston-kyselypalvelu-ogc-api/tekninen-kuvaus) for valid parameters.

## Best Practices

* **API Key Management**: Store your API key securely using `options(geofi_mml_api_key = "your_key")` rather than hardcoding it in scripts.
* **Large Datasets**: For collections like "`suo`", test with a small `limit` or `bbox` to estimate runtime before fetching all features.
* **CRS Selection**: Use EPSG:3067 (ETRS-TM35FIN) for Finnish data unless you need WGS84 (EPSG:4326) for compatibility with other systems.
* **Check Available Collections**: Always use `ogc_get_maastotietokanta_collections()` to verify collection names before downloading.

## Additional Resources

* Maastotietokanta Product Description
* Geographic Names Product Description
* NLS API Key Instructions

## Conclusion

The `geofi` package simplifies access to the Maastotietokanta and Geographic Names datasets, enabling spatial analysis of Finnish topographic and place name data. By handling API requests, pagination, and CRS transformations, it allows users to focus on data analysis and visualization. Try the examples above and explore the datasets to uncover insights about Finland's geography!
