---
title: "Making maps using geofi-package"
author: "Markus Kainu, Leo Lahti & Joona Lehtomäki"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Making maps using geofi-package}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


**Installation**

`geofi` is not yet in CRAN, but can be installed from Github using

```{r, eval = FALSE}
remotes::install_github("ropengov/geofi")
```



## Municipalities

```{r municipality_map, fig.height = 4, fig.width = 7}
library(geofi)
library(ggplot2)
library(dplyr)
library(janitor)
library(tidyr)

municipalities <- get_municipalities(year = 2021, scale = 4500)
ggplot(municipalities) + 
  geom_sf(aes(fill = as.integer(kunta)))
```

You can add the locations of central locality of each municipality with 

```{r}
ggplot(municipalities) + 
  geom_sf(aes(fill = as.integer(kunta))) +
  geom_sf(data = municipality_central_localities, shape = 5, color = "white")
```


## Zipcodes


```{r zipcode_map, fig.height = 4, fig.width = 7}
zipcodes <- get_zipcodes(year = 2021) 
ggplot(zipcodes) + 
  geom_sf(aes(fill = as.integer(posti_alue)))
``` 


You can add variables from Paavo data (Statistics Finland) in following manner.

```{r zipcode_with_statistics_finland, fig.height = 4, fig.width = 7}
library(pxweb)
# lets get all zipcodes and all variables
pxweb_query_list <- 
  list("Postinumeroalue"=c("*"),
                           "Tiedot"=c("*"))

# Download data 
px_data <- 
 pxweb_get(url = "http://pxnet2.stat.fi/PXWeb/api/v1/fi/Postinumeroalueittainen_avoin_tieto/2019/paavo_1_he_2019.px",
           query = pxweb_query_list)

# Convert to data.frame 
tk_data_zip <- as.data.frame(px_data, column.name.type = "text", variable.value.type = "text")

library(dplyr)
library(tidyr)
tk_data_zip2 <- tk_data_zip %>%
  as_tibble() %>% 
  setNames(c("region","indicator","value")) %>% 
  spread(indicator, value)
tk_zipdata <- janitor::clean_names(tk_data_zip2)
tk_zipdata$zipcode <- substr(x = tk_zipdata$region, start = 1, stop = 5)
tk_zipdata$zipname <- gsub("^.{6}", "", tk_zipdata$region)


# Lets join with spatial data and plot the area of each zipcode
zipcodes19 <- get_zipcodes(year = 2019) 
zipcodes_new <- left_join(zipcodes19, tk_zipdata, by = c("posti_alue" = "zipcode"))
ggplot(zipcodes_new) + 
  geom_sf(aes(fill = pinta_ala), color = alpha("white", 1/3))
```





## Joining municipality level data from Statistics Finland pxweb-api


```{r municipalities_with_data}
municipalities17 <- get_municipalities(year = 2017)

# pull municipality data from Statistics Finland
library(pxweb)
pxweb_query_list <-
  list("Alue 2019"=c("*"),
       "Tiedot"=c("M408","M411","M476","M391","M421","M478","M404","M410","M303","M297","M302","M44","M62","M70","M488","M486","M137","M140","M130","M162","M78","M485","M152","M72","M84","M106","M151","M499","M496","M495","M497","M498"),
       "Vuosi"=c("2017"))

px_data <-
  pxweb_get(url = "http://pxnet2.stat.fi/PXWeb/api/v1/fi/Kuntien_avainluvut/2019/kuntien_avainluvut_2019_aikasarja.px",
            query = pxweb_query_list)

# Convert to data.frame
tk_data <- as.data.frame(px_data, column.name.type = "text", variable.value.type = "text")
tk_data2 <- tk_data %>%
  rename(name = `Alue 2019`) %>%
  mutate(name = as.character(name),
         # Paste Tiedot and Vuosi
         Tiedot = paste(Tiedot, Vuosi)) %>%
  select(-Vuosi) %>%
  spread(Tiedot, `Kuntien avainluvut`) %>%
  as_tibble()
tk_data3 <- janitor::clean_names(tk_data2)

# Join with Statistics Finland attribute data
dat <- left_join(municipalities17, tk_data3)
dat[1:10,1:10]
```


## Aggregating municipality level data using internal municipality_key-files

```{r municipalities_map, fig.height = 4, fig.width = 7}
# Lets aggregate at maakunta level using column mk_name
mk_data <- dat %>% 
  group_by(maakunta_name_fi,maakunta_code) %>% 
  # And compute population at maakunta level
  summarise(vakiluku_2017 = sum(vakiluku_2017))

ggplot(mk_data) + 
  geom_sf(aes(fill = vakiluku_2017)) +
  geom_sf_text(aes(label = paste(maakunta_name_fi, "\n",vakiluku_2017)), 
               size = 3, 
               color = "white")

```


Or if you prefer aggregating at seutukunta-level you can

```{r vaalipiiri_map, fig.height = 4, fig.width = 7}
mk_data <- dat %>% 
  group_by(seutukunta_code,seutukunta_name_fi) %>% 
  # And compute population at maakunta level
  summarise(vakiluku_2017 = sum(vakiluku_2017))

ggplot(mk_data) + 
  geom_sf(aes(fill = vakiluku_2017)) +
  geom_sf_text(aes(label = paste(seutukunta_name_fi, "\n",vakiluku_2017)), 
               size = 3, 
               color = "white")

```


Or if you want the population at the national level aggregated up from municipality level you can: 

```{r, national_map, fig.height = 4, fig.width = 7}
mk_data <- dat %>% 
  summarise(vakiluku_2017 = sum(vakiluku_2017))

ggplot(mk_data) + 
  geom_sf(aes(fill = vakiluku_2017)) +
  geom_sf_text(aes(label = vakiluku_2017), 
               size = 3, 
               color = "white")

```

Helsinki at 1km by 1km as interactive leaflet map


```{r, population_grid_data2, fig.height = 4, fig.width = 7}
library(leaflet)
pop_grid <- get_population_grid(year = 2018, resolution = 1)

pop_grid_helsinki <- sf::st_transform(x = pop_grid %>% 
                                        filter(kunta == "091"), crs = "+proj=longlat +datum=WGS84")



leaflet(pop_grid_helsinki) %>% 
  addTiles() %>% 
  addPolygons(fillColor = ~colorQuantile("RdYlGn", vaesto)(vaesto),
              color = "coral", 
              weight = 1,
              smoothFactor = 0.5,
              opacity = 1.0,
              fillOpacity = 0.5,
    label = ~vaesto,
    highlightOptions = highlightOptions(color = "white", weight = 2,
      bringToFront = TRUE))
```



```{r geofacet2, fig.height = 6, fig.width = 7, eval = FALSE}

dat <- left_join(geofi::municipality_key_2019 %>% 
                   filter(maakunta_name_fi == "Ahvenanmaa") %>% 
                   select(-year),
                 px_data) %>% 
  group_by(maakunta_code, maakunta_name_fi,year) %>% 
  rename(code = kunta, name = kunta_name, population = value)

library(geofacet)
library(ggplot2)

ggplot(dat, aes(x = year, y = population/1000, group = name)) + 
  geom_line() + 
  facet_geo(facets = ~name, grid = grid_ahvenanmaa_2019, scales = "free_y") +
  theme(axis.text.x = element_text(size = 6)) +
  scale_x_discrete(breaks = seq.int(from = 1987, to = 2018, by = 5)) +
  labs(title = "Population 1987-2018", y = "population (1000)")
```
