---
title: "geofi tutorial"
author: "Markus Kainu, Leo Lahti & Joona Lehtomäki"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{geofi tutorial}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Installation

`geofi` is not yet in CRAN, but can be installed from Github using

```{r, eval = FALSE}
remotes::install_github("ropengov/geofi")
```


## Municipalities

```{r municipality_map, fig.height = 8}
library(geofi)
library(ggplot2)
library(dplyr)
library(janitor)
library(tidyr)


municipalities <- get_municipalities(year = 2019, scale = 4500)
ggplot(municipalities) + 
  geom_sf(aes(fill = as.integer(kunta)))
```

## Zipcodes


```{r zipcode_map, fig.height = 8}
zipcodes <- get_zipcodes(year = 2019) 
ggplot(zipcodes) + 
  geom_sf(aes(fill = as.integer(posti_alue)))
```


You can add variables from Paavo data (Statistics Finland) in following manner.

```{r zipcode_with_statistics_finland, fig.height = 8}
library(pxweb)
# lets get all zipcodes and all variables
pxweb_query_list <- 
  list("Postinumeroalue"=c("*"),
                           "Tiedot"=c("*"))

# Download data 
px_data <- 
 pxweb_get(url = "http://pxnet2.stat.fi/PXWeb/api/v1/fi/Postinumeroalueittainen_avoin_tieto/2019/paavo_1_he_2019.px",
           query = pxweb_query_list)

# Convert to data.frame 
tk_data_zip <- as.data.frame(px_data, column.name.type = "text", variable.value.type = "text")

library(dplyr)
library(tidyr)
tk_data_zip2 <- tk_data_zip %>%
  as_tibble() %>% 
  setNames(c("region","indicator","value")) %>% 
  spread(indicator, value)
tk_zipdata <- janitor::clean_names(tk_data_zip2)
tk_zipdata$zipcode <- substr(x = tk_zipdata$region, start = 1, stop = 5)
tk_zipdata$zipname <- gsub("^.{6}", "", tk_zipdata$region)


# Lets join with spatial data and plot the area of each zipcode
zipcodes_new <- left_join(zipcodes, tk_zipdata, by = c("posti_alue" = "zipcode"))
ggplot(zipcodes_new) + 
  geom_sf(aes(fill = pinta_ala), color = alpha("white", 1/3))
```





## Joining municipality level data from Statistics Finland


```{r municipalities_with_data}
municipalities17 <- get_municipalities(year = 2017)

# pull municipality data from Statistics Finland
library(pxweb)
pxweb_query_list <-
  list("Alue 2019"=c("SSS","020","005","009","010","016","018","019","035","043","046","047","049","050","051","052","060","061","062","065","069","071","072","074","075","076","077","078","079","081","082","086","111","090","091","097","098","099","102","103","105","106","108","109","139","140","142","143","145","146","153","148","149","151","152","165","167","169","170","171","172","176","177","178","179","181","182","186","202","204","205","208","211","213","214","216","217","218","224","226","230","231","232","233","235","236","239","240","320","241","322","244","245","249","250","256","257","260","261","263","265","271","272","273","275","276","280","284","285","286","287","288","290","291","295","297","300","301","304","305","312","316","317","318","398","399","400","407","402","403","405","408","410","416","417","418","420","421","422","423","425","426","444","430","433","434","435","436","438","440","441","475","478","480","481","483","484","489","491","494","495","498","499","500","503","504","505","508","507","529","531","535","536","538","541","543","545","560","561","562","563","564","309","576","577","578","445","580","581","599","583","854","584","588","592","593","595","598","601","604","607","608","609","611","638","614","615","616","619","620","623","624","625","626","630","631","635","636","678","710","680","681","683","684","686","687","689","691","694","697","698","700","702","704","707","729","732","734","736","790","738","739","740","742","743","746","747","748","791","749","751","753","755","758","759","761","762","765","766","768","771","777","778","781","783","831","832","833","834","837","844","845","846","848","849","850","851","853","857","858","859","886","887","889","890","892","893","895","785","905","908","911","092","915","918","921","922","924","925","927","931","934","935","936","941","946","976","977","980","981","989","992","MK01","MK02","MK04","MK05","MK06","MK07","MK08","MK09","MK10","MK11","MK12","MK13","MK14","MK15","MK16","MK17","MK18","MK19","MK21","SK011","SK014","SK015","SK016","SK021","SK022","SK023","SK024","SK025","SK041","SK043","SK044","SK051","SK052","SK053","SK061","SK063","SK064","SK068","SK069","SK071","SK081","SK082","SK091","SK093","SK101","SK103","SK105","SK111","SK112","SK113","SK114","SK115","SK122","SK124","SK125","SK131","SK132","SK133","SK134","SK135","SK138","SK141","SK142","SK144","SK146","SK151","SK152","SK153","SK154","SK161","SK162","SK171","SK173","SK174","SK175","SK176","SK177","SK178","SK181","SK182","SK191","SK192","SK193","SK194","SK196","SK197","SK211","SK212","SK213"),
       "Tiedot"=c("M408","M411","M476","M391","M421","M478","M404","M410","M303","M297","M302","M44","M62","M70","M488","M486","M137","M140","M130","M162","M78","M485","M152","M72","M84","M106","M151","M499","M496","M495","M497","M498"),
       "Vuosi"=c("2017"))

px_data <-
  pxweb_get(url = "http://pxnet2.stat.fi/PXWeb/api/v1/fi/Kuntien_avainluvut/2019/kuntien_avainluvut_2019_aikasarja.px",
            query = pxweb_query_list)

# Convert to data.frame
tk_data <- as.data.frame(px_data, column.name.type = "text", variable.value.type = "text")
tk_data2 <- tk_data %>%
  rename(name = `Alue 2019`) %>%
  mutate(name = as.character(name),
         # Paste Tiedot and Vuosi
         Tiedot = paste(Tiedot, Vuosi)) %>%
  select(-Vuosi) %>%
  spread(Tiedot, `Kuntien avainluvut`) %>%
  as_tibble()
tk_data3 <- janitor::clean_names(tk_data2)

# Join with Statistics Finland attribute data
dat <- left_join(municipalities17, tk_data3)
dat[1:10,1:10]
```


## Aggregating municipality level data using internal municipality_key-files

```{r municipalities_map, fig.height = 8}
# Lets join the municipality level data with internal municipality_key_2017-data
dat2 <- left_join(dat, municipality_key_2017, by = c("name" = "name_fi"))

# Lets aggregate at maakunta level using column mk_name
mk_data <- dat2 %>% 
  group_by(mk_name,mk_code) %>% 
  # And compute population at maakunta level
  summarise(vakiluku_2017 = sum(vakiluku_2017))

ggplot(mk_data) + 
  geom_sf(aes(fill = vakiluku_2017)) +
  geom_sf_text(aes(label = paste(mk_name, "\n",vakiluku_2017)), 
               size = 3, 
               color = "white")

```


Or if you prefer aggregating at vaalipiiri-level you can

```{r vaalipiiri_map, fig.height = 8}
mk_data <- dat2 %>% 
  group_by(va_name,va_code) %>% 
  # And compute population at maakunta level
  summarise(vakiluku_2017 = sum(vakiluku_2017))

ggplot(mk_data) + 
  geom_sf(aes(fill = vakiluku_2017)) +
  geom_sf_text(aes(label = paste(va_name, "\n",vakiluku_2017)), 
               size = 3, 
               color = "white")

```


Or if you want the population at the national level aggregated up from municipality level you can: 

```{r, national_map, fig.height = 8}
mk_data <- dat2 %>% 
  summarise(vakiluku_2017 = sum(vakiluku_2017))

ggplot(mk_data) + 
  geom_sf(aes(fill = vakiluku_2017)) +
  geom_sf_text(aes(label = vakiluku_2017), 
               size = 3, 
               color = "white")

```


## Interactive maps using `leaflet`-package



```{r}
library(leaflet)
library(geofi)
library(dplyr)

municipalities <- get_zipcodes(year = 2019)
municipalities_lonlat <- sf::st_transform(x = municipalities, crs = "+proj=longlat +datum=WGS84")

leaflet(municipalities_lonlat) %>% 
  addTiles() %>% 
  addPolygons(color = "coral", 
              weight = 1,
              smoothFactor = 0.5,
              opacity = 1.0,
              fillOpacity = 0.5,
    label = ~nimi,
    highlightOptions = highlightOptions(color = "white", weight = 2,
      bringToFront = TRUE))

```

