<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{An R Markdown Vignette made with knitr}
-->

<!--
Joissakin esimerkeissa tarvittavat riippuvuudet 
  <a href="http://trac.osgeo.org/gdal/wiki/DownloadSource">GDAL</a>, 
  <a href="http://freeglut.sourceforge.net/">freeglut</a>, 
	<a href="http://xmlsoft.org/downloads.html">XML</a>, 
	<a href="http://trac.osgeo.org/geos">GEOS</a> ja 
	<a href="http://trac.osgeo.org/proj">PROJ.4</a>. 
-->

```{r knit, eval=FALSE, echo=FALSE}
# How to knit this document into .md and .html
library(knitr)
opts_knit$set(base.dir = "vignettes") # Change the base dir where to save figures
knit(input="vignettes/gisfin_tutorial.Rmd", 
     output="vignettes/gisfin_tutorial.md")
knit2html(input="vignettes/gisfin_tutorial.md", 
          output="vignettes/gisfin_tutorial.html", 
          options=c("use_xhtml","smartypants","mathjax","highlight_code"))
```

# gisfin - tutorial

This R package provides tools to access open spatial data in Finland
as part of the [rOpenGov](http://ropengov.github.io) project.

For contact information and source code, see the 
[github page](https://github.com/rOpenGov/gisfin)

## Available data sources

[Helsinki region district maps](#aluejakokartat) (Helsingin seudun 
aluejakokartat)

+ Aluejakokartat: kunta, pien-, suur-, tilastoalueet (Helsinki region district 
maps)
+ Äänestysaluejako: (Helsinki region election district maps)
+ Source: [Helsingin kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/)

[Helsinki spatial data](#hel-spatial) (Helsingin seudun avoimia 
paikkatietoaineistoja)

+ Seutukartta (Helsinki Region Maps)
+ Helsingin piirijako (District Division of the City of Helsinki)
+ Seudullinen osoiteluettelo (Regional Address List)
+ Helsingin osoiteluettelo (Register of Addresses of the City of Helsinki)
+ Rakennusrekisterin ote (Helsinki building registry)
+ Source: [Helsingin kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/)

[National Land Survey data](#maanmittauslaitos) (Maanmittauslaitoksen avointa 
dataa)

+ Yleiskartat: kunta-, maakuntarajat (municipality and province borders)
+ Source: [Maanmittauslaitos (MML)](http://www.maanmittauslaitos.fi/avoindata)

[Geocoding](#geocoding)

+ [OKF.fi Geocoding API Test Console](http://data.okf.fi/console/)
+ [OpenStreetMap Nominatim](http://wiki.openstreetmap.org/wiki/Nominatim_usage_policy)
+ [Google](http://code.google.com/apis/maps/documentation/geocoding/)

[IP address geographic coordinates](#ip) 

+ [Data Science Toolkit](http://www.datasciencetoolkit.org/)

[Statistics Finland geospatial data](#geostatfi) (Tilastokeskuksen paikkatietoaineistoja)

+ Väestöruutuaineisto (Population grid)
+ Tuotanto- ja teollisuuslaitokset (Production and industrial facilities)
+ Oppilaitokset (Educational institutions)
+ Tieliikenneonnettomuudet (Road accidents)
+ Source: [Statistics Finland](http://www.stat.fi/tup/rajapintapalvelut/index_en.html)

List of potential data sources to be added to the package can be found 
[here](https://github.com/rOpenGov/gisfin/blob/master/vignettes/todo-datasets.md).

## Installation

### Requirements

The gisfin package uses the 
[rgdal](http://cran.r-project.org/web/packages/rgdal/index.html) package, which 
depends on [GDAL](http://www.gdal.org/) (Geospatial Data Abstraction Library). 
Some rgdal installation tips for various platforms are listed below. If you 
encounter problems, please contact us by email: louhos@googlegroups.com.

#### Windows

Install binaries from [CRAN](http://cran.r-project.org/web/packages/rgdal/index.html)

#### OSX

It is preferable to NOT use the [CRAN binaries](http://cran.r-project.org/web/packages/rgdal/index.html). Instead, you should check [KyngChaos Wiki](http://www.kyngchaos.com/software/frameworks)
and install the latest GDAL Complete (v. 1.11 in July 2014) and then install the 
corresponding rgdal package from the same page (v. 0.8.16 in July 2014). 

#### Linux 

Install the following packages through your distribution's package manager

__Ubuntu/Debian__

```
sudo apt-get -y install libgdal1-dev libproj-dev
```

__Fedora__

```
sudo yum -y install gdal-devel proj-devel
```

__openSUSE__

```
sudo zypper --non-interactive in libgdal-devel libproj-devel
```

### Installing the package

Release version for general users:

```{r install, eval=FALSE}
install.packages("gisfin")
```

Development version for developers:

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("gisfin", "ropengov")
```

Load package.

```{r load, message=FALSE, warning=FALSE, results='hide'}
library(gisfin)
```

----

## <a name="aluejakokartat"></a>Helsinki region district maps

Helsinki region district maps (Helsingin seudun aluejakokartat) from 
[Helsingin kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/).

List available maps with `get_helsinki_aluejakokartat()`.

```{r aluejako_list, message=FALSE}
get_helsinki_aluejakokartat()
```

Below the 'suuralue' districts is used for plotting examples with `spplot()` and 
[ggplot2](http://ggplot2.org/). The other district types can be plotted 
similarly.

### Plot with spplot

Retrieve 'suuralue' spatial object with `get_helsinki_aluejakokartat()` and plot 
with `spplot()`.

```{r hkk-suuralue1, message=FALSE, warning=FALSE, results='hide'}
sp.suuralue <- get_helsinki_aluejakokartat(map.specifier="suuralue")
spplot(sp.suuralue, zcol="Name")
```

Function `generate_map_colours()` allows nice region colouring separable 
adjacent regions. This is used here with the `rainbow()` colour scale to plot 
the regions with `spplot()`.

```{r hkk-suuralue2, message=FALSE, warning=FALSE, results='hide'}
sp.suuralue@data$COL <- factor(generate_map_colours(sp=sp.suuralue))
spplot(sp.suuralue, zcol="COL", 
       col.regions=rainbow(length(levels(sp.suuralue@data$COL))), 
       colorkey=FALSE)
```

### Plot with ggplot2

Use the 'sp.suuralue' retrieved above, and retrieve also the center points of 
the districts. Use `sp2df()` function to tranform the spatial objects into data 
frames. Plot with [ggplot2](http://ggplot2.org/), using blank map theme with 
`get_theme_map()`. 

```{r hkk-suuralue3, message=FALSE, warning=FALSE, results='hide'}
# Retrieve center points
sp.suuralue.piste <- get_helsinki_aluejakokartat(map.specifier="suuralue_piste")
# Get data frames
df.suuralue <- sp2df(sp.suuralue, "Name")
df.suuralue.piste <- sp2df(sp.suuralue.piste, "Name")
# Set map theme
library(ggplot2)
theme_set(get_theme_map())
# Plot regions, add labels using the points data
ggplot(df.suuralue, aes(x=long, y=lat)) + geom_polygon(aes(fill=COL, 
                                                           group=Name)) + 
  geom_text(data=df.suuralue.piste, aes(label=Name)) + 
  theme(legend.position="none")
```

### Plot election districts

Retrieve and plot äänetysaluejako (election districts) with 
`get_helsinki_aluejakokartat()` and `spplot()`, use colours to separate 
municipalities.

```{r hkk-aanestysalue, message=FALSE, warning=FALSE, results='hide'}
sp.aanestys <- get_helsinki_aluejakokartat(map.specifier="aanestysalue")
spplot(sp.aanestys, zcol="KUNTA", 
       col.regions=rainbow(length(levels(sp.aanestys@data$KUNTA))), 
       colorkey=FALSE)

```

----

## <a name="hel-spatial"></a>Helsinki spatial data

Other Helsinki region spatial data from 
[Helsingin Kaupungin Kiinteistövirasto (HKK)](http://ptp.hel.fi/avoindata/).

List available spatial data with `get_helsinki_spatial()`.

```{r hki_spatial_list, message=FALSE, warning=FALSE, results='hide'}
get_helsinki_spatial()
```

Retrieve municipality map for the larger Helsinki region with 
`get_helsinki_spatial()` and transform coordinates with `sp::spTransform()`.

```{r peruspiiri, message=FALSE, warning=FALSE, results='hide'}
sp.piiri <- get_helsinki_spatial(map.type="piirijako", 
                                 map.specifier="ALUEJAKO_PERUSPIIRI")
# Check current coordinates
sp.piiri@proj4string
# Transform coordinates to WGS84
sp.piiri <- sp::spTransform(sp.piiri, CRS("+proj=longlat +datum=WGS84"))
```

----

## <a name="maanmittauslaitos"></a>National Land Survey Finland

Spatial data from [National Land Survey Finland](http://www.maanmittauslaitos.fi/en/opendata) 
(Maanmittauslaitos, MML). These data are preprocessed into RData format, see 
details [here](https://github.com/avoindata/mml).

List available data sets with `list_mml_datasets()`.

```{r MML_list, message=FALSE} 
list_mml_datasets()
```

Retrieve regional borders for Finland with `get_mml()`.

```{r MML_regions, message=FALSE, warning=FALSE} 
# Get a specific map
sp.mml <- get_mml(map.id="Yleiskartta-4500", data.id="HallintoAlue")
# Investigate available variables in this map
head(as.data.frame(sp.mml))
```

Plot municipalities (kunnat) with `spplot()`, using colours from 
`generate_map_colours()`.


```{r MML_municipality, message=FALSE, warning=FALSE}
# Get region colouring for municipalities
sp.mml@data$COL <- factor(generate_map_colours(sp.mml))
# Plot the shape file, colour municipalities
spplot(sp.mml, zcol="COL", col.regions=rainbow(length(levels(sp.mml@data$COL))), 
       colorkey=FALSE)
```

----

## <a name="geocoding"></a>Geocoding

Get geocodes for given location (address etc.) using one of the available 
services. Please read carefully the usage policies for the different services:

+ [OKF.fi Geocoding API Test Console](http://data.okf.fi/console/)
+ [OpenStreetMap Nominatim](http://wiki.openstreetmap.org/wiki/Nominatim_usage_policy)
+ [Google](http://code.google.com/apis/maps/documentation/geocoding/)

The function `get_geocode()` returns both latitude and longitude for the first 
hit, and the raw output (varies depending on the service used).

Warning! The geocode results may vary between sources, use with care!


```{r geocoding, message=FALSE}
gc1 <- get_geocode("Mannerheimintie 100, Helsinki", service="okf")
unlist(gc1[1:2])
gc2 <- get_geocode("Mannerheimintie 100, Helsinki", service="openstreetmap")
unlist(gc2[1:2])
gc3 <- get_geocode("Mannerheimintie 100, Helsinki", service="google")
unlist(gc3[1:2])
```

----

## <a name="ip"></a>IP Location

Geographic coordinates for a given IP-address from
[Data Science Toolkit](http://www.datasciencetoolkit.org/):

```{r iplocation, message=FALSE, eval=TRUE}
ip_location("137.224.252.10")
```

----

## <a name="geostatfi"></a>Statistics Finland geospatial data

Geospatial data provided by 
[Statistics Finland](http://www.stat.fi/tup/rajapintapalvelut/inspire_en.html).

Retrieve a list of the available data sets for population density. In case the 
service is unreachable, `character(0)` is returned.

```{r population-density-list, message=FALSE}
geoStatFi <- GeoStatFi()
layers <- geoStatFi$listPopulationLayers()
if (length(layers) > 0) layers
```

Get population density in year 2005 on a 5 km x 5 km grid, convert to 
RasterStack object and plot.

```{r population-density-plot, message=FALSE, warning=FALSE, results='hide'}
population <- geoStatFi$getPopulation(layers[11])
if (length(population) > 0) {
  x <- geoStatFi$getRaster(population)
  plot(x[["vaesto"]])
}
```

----

### Citation

**Citing the data:** See `help()` to get citation information for each data 
source individually.

**Citing the R package:**

```{r citation, comment=NA}
citation("gisfin")
```


### Session info


This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```

