## Finnish geographic data in R 

This page shows how use standard data sets (municipality borders etc.)
from Land Survey Finland.

### Available data sets

To list the latest MML data sets that are directly available in R
through the sorvi package in
[avoindata/mml](http://www.datavaalit.fi/storage/avoindata/mml/rdata/),
type in R the following. Note that [2012 data
sets](http://www.datavaalit.fi/storage/avoindata/mml/rdata/2012/) in
RData format are available but not shown in the above automated
listing.


```{r MMLShape, message=FALSE, fig.width=8, fig.height=8, warning=FALSE} 
library(sorvi, quietly = TRUE)
list_mml_datasets()
```

### Municipality borders

```{r MMLShape2, message=FALSE, fig.width=8, fig.height=8, warning=FALSE} 

# Get a specific map (see list_mml_datasets() for options)
map.id <- "Yleiskartta-4500"
data.id <- "HallintoAlue"
sp <- LoadMML(map.id, data.id) 

# Investigate available variables in this map
varnames <- names(sp)
print(varnames)
head(as.data.frame(sp))

# Visualize the shape file
p <- PlotShape(sp, "Maakunta", plot = FALSE)
p

# Print the plot to PDF file and open from command line
# (done this ways as there seem to be some problem with spplot PNG file outputs currently)
pdf("Maakunta.pdf")
print(p)
dev.off()
system("evince Maakunta.pdf &")
```


### Shape data files 

Assuming you have downloaded a shape file f, adjust these
commands to read in, modify, and visualize the data:

```{r MMLShape3, message=FALSE, eval = FALSE} 

# Read and preprocess shape file
sp <- maptools::readShapePoly(f) # read the shapefile
sp <- sorvi::PreprocessShapeMML(sp) # handle special characters etc.

# Add some data fields to plot
sp$example <- rnorm(nrow(sp))
varname <- "example"
int <- max(abs(sp[[varname]]))

# Visualize the shape file
varname <- names(sp)[[2]] # Which layer to show? 
p <- PlotShape(sp, varname, type = "twoway", main = "Title", at = seq(0 - int, 0 + int, length = 11), plot = FALSE)
print(p)
```


### XML data files

To get data in R, get the URL for your desired data in
[Kapsi](http://kartat.kapsi.fi/), and provide it as in this
example. We have noticed problems with certain XML files. We welcome
all bug reports and improvements to these tools and instructions, just
[contact us](contact).


```{r MMLXML, message=FALSE, fig.width=8, fig.height=8} 
library(sorvi)

# Define url from http://kartat.kapsi.fi
url <- "http://kartat.kapsi.fi/files/kuntajako/kuntajako_4500k/etrs89/gml/TietoaKuntajaosta_2013_4500k.zip"

# Define temporary data directory
data.dir <- gsub(" ", "-", paste("tmp.data.", date(), sep = ""))
system(paste("mkdir ", data.dir))
local.zip <- paste(data.dir, "/tmp.zip", sep = "")

# Download the file:
download.file(url, destfile = local.zip)

# Unzip the downloaded zip file
data.dir <- file.path(data.dir)
# system(paste('unzip', local.zip))
unzip(local.zip, exdir = data.dir)

# Remove TEMP file
unlink("TEMP", recursive=T)

# In the above we have this file:
my.file <- "TietoaKuntajaosta_2013_4500k/SuomenKuntajako_2013_4500k.xml"

# List available map layers
library(rgdal)
boundaries <- readOGR(my.file, layer = ogrListLayers(my.file)[[2]])

# Plotting
p <- PlotShape(boundaries, names(boundaries)[[1]], plot = FALSE)
print(p)
```




### Municipality boundaries for land areas

This is often asked for. 

For the 2012 municipality boundaries, you can use this version:

```{r MML-maa, message=FALSE}
sp <- LoadData("kuntarajat.maa.shp") 
```

The 2013 version without sea regions is not yet readily available in R, but you can follow [these instructions](https://github.com/kansanmuisti/datavaalit/wiki/kuntarajat-leikkatuna-suomen-rantaviivalla) to cut off the sea regions from the overall map and carry out [R conversion](https://github.com/louhos/preprocessing/blob/master/MML-maa-alueet.R). Contributions welcome! 


### Conversions from MML/Kapsi to RData (for developers)

RData conversions are documented [here](https://github.com/avoindata/mml)


### TODO

See [TODO](mml.url)


### Background

Finnish geographic information (GIS) is available [through Land Survey
Finland (MML)](https://tiedostopalvelu.maanmittauslaitos.fi/tp/kartta)
in XML/GML format, but this requires a separate query. The same data
is directly available [through Kapsi](http://kartat.kapsi.fi/), and we
have further [converted](https://github.com/avoindata/mml) the key
data sets into light-weight RData format that we distribute
[here](http://www.datavaalit.fi/storage/avoindata/mml/).

### Version information
 
```{r versiotiedot}
sessionInfo()
```


