<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{An R Markdown Vignette made with knitr}
-->

Finland GIS R tools
===========

This is an [rOpenGov](http://ropengov.github.com/helsinki) R package
providing tools for Finnish GIS data.


### Installation

Release version for general users:

```{r install, eval=FALSE}
install.packages("gisfi")
library(gisfi)
```

Development version for developers:

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("gisfi", "ropengov")
library(gisfi)
```

Further installation and development instructions at the [home
page](http://ropengov.github.com/gisfi).


### Election districts (Aanestysaluejako)

Aanestysaluejako data set includes the elections districts for the cities of the capital region (Helsinki, Espoo, Kauniainen and Vantaa) in MapInfo-format. CRS is Finland KKJ2 (EPSG: 2392). 

```{r hkk-electiondistricts, message=FALSE}
# Aanestysaluejako
sp <- GetElectionDistrictsHelsinki()
```


### Statistics districts (Tilastoalue)

```{r hkk-tilastoalue, message=FALSE}
id <- "tilastoalue"
res <- GetDistrictBoundariesHelsinki(id); 
sp <- res$shape; 
df <- res$data
head(df)
PlotShape(sp, "Name")
```

### Municipalities (Kunta)

```{r hkk-kunta, message=FALSE}
id <- "kunta"
res <- GetDistrictBoundariesHelsinki(id); 
sp <- res$shape; 
df <- res$data
head(df)
PlotShape(sp, "Name")
```

### Major districts (Suuralue)

```{r hkk-suuralue, message=FALSE}
id <- "suuralue"
res <- GetDistrictBoundariesHelsinki(id); 
sp <- res$shape; 
df <- res$data
head(df)
PlotShape(sp, "Name")
```

### Minor districts (Pienalue)

```{r hkk-pienalue, message=FALSE}
id <- "pienalue"
res <- GetDistrictBoundariesHelsinki(id); 
sp <- res$shape; 
df <- res$data
head(df)
PlotShape(sp, "Name")
```



### Licensing and Citations

This work can be freely used, modified and distributed under the
[Two-clause FreeBSD
license](http://en.wikipedia.org/wiki/BSD\_licenses). Cite this R
package and and the appropriate data provider, including a url
link. Kindly cite the R package as 'Leo Lahti, Juuso Parkkinen ja
Joona Lehtomäki (2013). gisfi R package. URL:
http://ropengov.github.io/gisfi'.


### Session info


This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```

